`r params$doc_title`

### 2022-11-22 scripts/MSc.Rmd
outsourced fns

library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(ggpubr)
library(scales)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(TwoSampleMR)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
source("MR_lasso.R")  # Slob&Burgess 2020
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

@@

```{r calculate-explained-variance}
explained_variance_numeric <- function(eaf, beta) {
  # eaf = allele frequency
  # beta = additive genetic effect
  explained_variance_numeric <- 2*eaf *(1-eaf) * beta^2
  return(explained_variance_numeric)
}

explained_variance_numeric2 <- function(maf, beta, se_beta, samplesize) {
  ### Teslovich Nature 2010
  # 1) maf = minor allele frequency
  # 2) beta = additive genetic effect
  # 3) se_beta = standard error of beta estimate
  # 4) samplesize = samplesize of study
  maf <- ifelse(maf>0.5, 1-maf, maf)

  explained_variance_numeric <- (2*beta^2*maf*(1-maf))/ 
    ( 2*beta^2*maf*(1-maf) + (se_beta^2)*2*samplesize*maf*(1-maf) )
  return(explained_variance_numeric)
}

explained_variance_binary <- function(PA,RR1,RR2,K) {
  # from So et al 2011
  # Calculates the variance in liability explained by a single biallelic loci.
  # PA: allele frequency of the risk allele (denoted A)
  # RR1: relative risk of Aa (one risk allele) compared to aa (no risk allele)
  # RR2: relative risk of AA (two risk alleles) compared to aa (no risk allele)
  # K:  overall probability of disease in population
  # Returns the variance explained (Vg) and the mean liability for each genotype (the overall liability is normalized to mean 0 and variance 1)
  Paa = (1-PA)^2
  PAa = 2*PA*(1-PA)
  PAA = PA^2
  muaa=0
  faa= K/(Paa + PAa*RR1 + PAA*RR2)
  fAa= RR1*faa
  fAA= RR2*faa 
  T = qnorm(1-faa) 
  muAa = T-qnorm(1-fAa)
  muAA = T-qnorm(1-fAA)
  mean.all= PAa*muAa+ PAA*muAA
  Vg= Paa*(muaa-mean.all)^2 + PAa*(muAa-mean.all)^2+ PAA*(muAA-mean.all)^2
  actual.Vg =  Vg/(1+Vg) 
  VR = 1-actual.Vg 
  actual.T = Paa*sqrt(VR)*qnorm(1-faa) + PAa*sqrt(VR)*qnorm(1-fAa) + PAA*sqrt(VR)*qnorm(1-fAA)
  actual.muaa = actual.T - sqrt(VR) * qnorm(1-faa)
  actual.muAa = actual.T - sqrt(VR) * qnorm(1-fAa)
  actual.muAA = actual.T - sqrt(VR) * qnorm(1-fAA)

  res <- list(Vg=actual.Vg,muaa=actual.muaa, muAa = actual.muAa, muAA=actual.muAA)
  res
}

## calculate explained variance
my_data_harm <- my_data_harm %>% 
  mutate(explained_variance_BMI = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi), 
         explained_variance_BMI2 = explained_variance_numeric2(maf=eaf.bmi, 
                                                           beta=beta.bmi, 
                                                           se_beta=se.bmi, 
                                                           samplesize=n[length(n)]

         ),
         explained_variance_GSD = explained_variance_binary(PA = eaf.gsd, 
                                                            RR1 = exp(beta.gsd), 
                                                            RR2 = exp(beta.gsd)^2, 
                                                            K = 0.1  # assumed prevalence in population
                                                            )$Vg,
         explained_variance_GBC = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc), 
                                                            RR2 = exp(beta.gbc)^2, 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg
         )
```

@@

ggpubr::ggarrange(fig01b.update.arrange, fig02.update.arrange,
                  gfigure01_EFFECT_SIZE_strong_VARIANTS_common.arrange, figure01_EFFECT_SIZE_strong_VARIANTS_rare.arrange,
                  figure01_EFFECT_SIZE_weak_VARIANTS_common.arrange, figure01_EFFECT_SIZE_weak_VARIANTS_rare.arrange, 
                  ncol=2, nrow=3, 
                  heights = c(2, 1.5, 1.5),
                  legend = "bottom", common.legend = TRUE, align = "hv"
                  ) -> final_plot1.arrange

figure01_EFFECT_SIZE_strong_VARIANTS_rare.arrange+
  scale_y_continuous(name="", sec.axis = sec_axis(~./scale_factor, name=""), limits = c(0,2))

    scale_y_continuous(
      # Features of the first axis
      name = "% variance explained by IVs",
      # Add a second axis and specify its features
      sec.axis = sec_axis(~./scale_factor, name="Number of IVs"),
      limits = c(0,2)
    ) +

@@

saveRDS(est, paste0("../output/Rdata/", Sys.Date(), "_est.rds"))
```

@@

```{r simulatesimulatesimulate, warning=FALSE}
# try starting with n=50.000 -> selected 9 IVs
set.seed(1234)

@@

 # simulate beta values from normal_distribution
 for (k in 1:nsim) {
   sim_dat <- sim_dat %>% 
     mutate(beta.bmi_sim_norm = purrr::map2_dbl(.x = beta.bmi, .y = se_update_BMI, .f = ~rnorm(n = 1, mean = .x, sd = .y))) %>% 
     mutate(beta.gbc_sim_norm = purrr::map2_dbl(.x = beta.gbc, .y = se_update_GBC, .f = ~rnorm(n = 1, mean = .x, sd = .y)))

   # PERFORM MR_analysis
   mr.obj = MendelianRandomization::mr_input(
         bx = sim_dat$beta.bmi_sim_norm, 
         bxse = sim_dat$se_update_BMI, 
         # outcome
         by = sim_dat$beta.gbc_sim_norm, 
         byse = sim_dat$se_update_GBC, 
         snps = sim_dat$SNP,
         exposure = "Body mass index",
         outcome = "Gallbladder Cancer"
       ) 
   # CONMIX (as reference)
   res = MendelianRandomization::mr_conmix(mr.obj)
   colname_theta <- paste0("theta_sim_", format(n[i], scientific=FALSE))
   colname_theta_se <- paste0("theta_se_sim_", format(n[i], scientific=FALSE))
   ## store results
   est_sim[k, colname_theta] <- res$Estimate
   CIlength = res$CIUpper-res$CILower
   if (length(CIlength)>1) print("conmix multimodal")
   est_sim[k, colname_theta_se] <- sum(CIlength)/1.96/2  ## Caution: this may be problematic 
   rm(res)
 }
 rm(se_update_BMI, se_update_GBC)
 print(paste0("|||-----------------------Run finished for sample size: ", format(n[i], scientific=FALSE), " -----------------------|||"))
}

>>>>

library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(ggthemes)
library(ggpubr)
library(scales)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(TwoSampleMR)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
source("MR_lasso.R")  # Slob&Burgess 2020
source("PVE_calculation_fns.R") 
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

@@

```{r calculate-explained-variance}
# outsourced to "./scripts/PVE_calculation_fns.R"
```

@@

ggpubr::ggarrange(fig01b.update.arrange, fig02.update.arrange,
                  figure01_EFFECT_SIZE_strong_VARIANTS_common.arrange, figure01_EFFECT_SIZE_strong_VARIANTS_rare.arrange,
                  figure01_EFFECT_SIZE_weak_VARIANTS_common.arrange, figure01_EFFECT_SIZE_weak_VARIANTS_rare.arrange, 
                  ncol=2, nrow=3, 
                  heights = c(2, 1.5, 1.5),
                  legend = "bottom", common.legend = TRUE, align = "hv"
                  ) -> final_plot1.arrange

figure01_EFFECT_SIZE_strong_VARIANTS_common.arrange+
  # delete y-labels
  scale_y_continuous(name="% variance explained by IVs", sec.axis = sec_axis(~./scale_factor, name=""), limits = c(0,2)) -> figure01_EFFECT_SIZE_strong_VARIANTS_common.arrange

figure01_EFFECT_SIZE_strong_VARIANTS_rare.arrange+
  # delete y-labels
  scale_y_continuous(name="", sec.axis = sec_axis(~./scale_factor, name=""), limits = c(0,2)) -> figure01_EFFECT_SIZE_strong_VARIANTS_rare.arrange

figure01_EFFECT_SIZE_weak_VARIANTS_common.arrange+
  # delete y-labels
  scale_y_continuous(name="",sec.axis = sec_axis(~./scale_factor, name=""), limits = c(0,2)) -> figure01_EFFECT_SIZE_weak_VARIANTS_common.arrange

figure01_EFFECT_SIZE_weak_VARIANTS_rare.arrange+
  # delete y-labels
  scale_y_continuous(name="", sec.axis = sec_axis(~./scale_factor, name="Number of IVs"), limits = c(0,2)) -> figure01_EFFECT_SIZE_weak_VARIANTS_rare.arrange

@@

saveRDS(est, paste0("../output/Rdata/", Sys.Date(), "_est.rds"))

@@

```{r simulatesimulatesimulate, warning=FALSE}

# data FOR simulation in simulation script
saveRDS(my_data_harm, paste0("../output/Rdata/", Sys.Date(), "_my_data_harm.rds"))
sim_vars <- list(n=n, colnames_pval=colnames_pval)
saveRDS(sim_vars, paste0("../output/Rdata/", Sys.Date(), "_sim_vars.rds"))

# try starting with n=50.000 -> selected 9 IVs
set.seed(1234)

@@

 # simulate beta values from normal_distribution # -----------------------------------------------------
 for (k in 1:nsim) {
   sim_dat <- sim_dat %>% 
     mutate(beta.bmi_sim_norm = purrr::map2_dbl(.x = beta.bmi, .y = se_update_BMI, .f = ~rnorm(n = 1, mean = .x, sd = .y))) %>% 
     mutate(beta.gbc_sim_norm = purrr::map2_dbl(.x = beta.gbc, .y = se_update_GBC, .f = ~rnorm(n = 1, mean = .x, sd = .y))) %>% 
     mutate(#explained_variance_X_sim = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi), 
         explained_variance_X2_sim = explained_variance_numeric2(maf=eaf.bmi, 
                                                           beta=beta.bmi_sim_norm, 
                                                           se_beta=se_update_BMI, 
                                                           samplesize=n[i]
         ),
        # explained_variance_GSD = explained_variance_binary(PA = eaf.gsd, 
        #                                                    RR1 = exp(beta.gsd), 
        #                                                    RR2 = exp(beta.gsd)^2, 
        #                                                    K = 0.1  # assumed prevalence in population
        #                                                    )$Vg,
         explained_variance_Y_sim = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc_sim_norm), 
                                                            RR2 = exp(beta.gbc_sim_norm)^2, 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg
         )

   # PERFORM MR_analysis
   mr.obj = MendelianRandomization::mr_input(
         bx = sim_dat$beta.bmi_sim_norm, 
         bxse = sim_dat$se_update_BMI, 
         # outcome
         by = sim_dat$beta.gbc_sim_norm, 
         byse = sim_dat$se_update_GBC, 
         snps = sim_dat$SNP,
         exposure = "Body mass index",
         outcome = "Gallbladder Cancer"
       ) 
   # CONMIX (as reference)
   res = MendelianRandomization::mr_conmix(mr.obj)
   colname_theta <- paste0("theta_sim_", format(n[i], scientific=FALSE))
   colname_theta_se <- paste0("theta_se_sim_", format(n[i], scientific=FALSE))
   colname_pve_X <- paste0("pve_X_sim_", format(n[i], scientific=FALSE))
   colname_pve_Y <- paste0("pve_Y_sim_", format(n[i], scientific=FALSE))

   ## store results
   est_sim[k, colname_theta] <- res$Estimate
   CIlength = res$CIUpper-res$CILower
   if (length(CIlength)>1) print("conmix multimodal")
   est_sim[k, colname_theta_se] <- sum(CIlength)/1.96/2  ## Caution: this may be problematic 
   # add PVE cols
   est_sim[k, colname_pve_X] <- sum(sim_dat$explained_variance_X2_sim)
   est_sim[k, colname_pve_Y] <- sum(sim_dat$explained_variance_Y_sim)

   rm(res)
 }
 rm(se_update_BMI, se_update_GBC)
 print(paste0("|||-----------------------Run finished for sample size: ", format(n[i], scientific=FALSE), " -----------------------|||"))
}
